---
title: "Final Project"
author: "Dominik Graetz"
date: "2023-02-25"
output: 
  html_document:
    toc: true
    toc_float: true
---

# Initializing, Data loading
This code takes forever, so I executed it only once and saved the data as an RDS
```{r setup, include=FALSE}

knitr::opts_chunk$set(cache = TRUE)
#library(WHO)
#y <- get_data("LIFE_0000000031")
#saveRDS(y, "data/WHOData_downloaded_1_30_23.RDS")
```


```{r}
options(scipen = 999)
library(tidyverse)
library(forcats)
library(janitor)
library(openxlsx)
library(ggpubr)
library(countrycode)

data <- readRDS("data/WHOData_downloaded_1_30_23.RDS")
countries <- read.csv("data/country_names.csv")

data <- data %>%
  mutate(agegroup = factor(agegroup),
         agegroup = factor(agegroup, levels = c(levels(agegroup)[1:2], levels(agegroup)[11], levels(agegroup)[3:10], levels(agegroup)[12:19])),
         prob = (value/100000)*100) %>%
  filter(!is.na(country),
         !is.na(agegroup))

data <- left_join(data, countries %>% select(DisplayString, ISO), by = c("country" = "DisplayString"))

```

# Exploratory Plot

```{r}

#useless plot, plotting individual survival traces for all countries for 2019
ggplot(data[data$year == 2019 & data$sex != "Both sexes",])+
  geom_line(aes(x = agegroup, y = prob, group = interaction(country, sex), color = sex), alpha = .2)+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1))
```

# Research Question 1: Did the survival rate increase over the past samples?
## Attempt 1, very simple
```{r}
#May be worth splitting by Region.
RQ1 <- data %>%
  filter(sex == "Both sexes") %>%
  group_by(agegroup, year) %>%
  summarize(M = mean(prob),
            SD = sd(prob),
            N = n(),
            SE = SD/sqrt(N),
            CI = SE*qt(.025, N-1, lower.tail = FALSE))

ggplot(RQ1) +
  geom_line(aes(x  = agegroup, y = M, group = year, color = year))+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1))

```

## Errorbars!

```{r}
pos <- position_dodge(width = .6)
ggplot(RQ1, aes(x  = agegroup, y = M, group = year, color = year)) +
  geom_line(position = pos, lwd = 1)+
  scale_color_continuous(breaks = c(2000, 2005, 2010, 2015, 2019))+
  geom_errorbar(aes(ymin = M - CI, ymax = M + CI), width = 0, position = pos, lwd = 1)+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## That's a nice first version!
Thinking about removing the color bar and just leaving the year as text. 

```{r}
RQ1_text <- RQ1[RQ1$agegroup == "85+ years",]
RQ1_text$y <- RQ1_text$M - mean(RQ1_text$M)
RQ1_text$y <- RQ1_text$y*1.5 + RQ1_text$M

pos <- position_dodge(width = .6)
ggplot(RQ1, aes(x  = agegroup, y = M, group = year, color = year)) +
  geom_line(position = pos, lwd = 1)+
  geom_errorbar(aes(ymin = M - CI, ymax = M + CI), width = 0, position = pos, lwd = 1)+
  geom_text(data = RQ1_text, aes(label = year, y = y), position = pos, hjust = -0.4, fontface = "bold")+
  labs(title = "Younger generations live longer.",
       subtitle = paste0("Data averaged across ", RQ1$N[1], " countries"),
       x = "Age Group",
       y = "Percentage alive",
       color = "Sample Year",
       caption = "Data from the WHO\nErrorbars represent confidence intervals.")+
  scale_y_continuous(limits = c(0, 100), 
                     labels = function(y) paste0(y, " %"))+
  scale_color_viridis_c(option = "plasma", 
                        guide = guide_colourbar(direction = "horizontal", 
                                                title.position = "top"), 
                        end = .9,
                        breaks = c(2000, 2005, 2010, 2015, 2019))+
  expand_limits(x = c(0, 22))+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = c(0.1, 0.2),
        legend.key.width = unit(1, 'cm'),
        legend.key.height = unit(0.4, 'cm'),
        legend.justification = "left")

```

## By region
Probably worth it, there are some interesting differences.

```{r}
#BY REGION

RQ1_text <- RQ1[RQ1$agegroup == "85+ years",]
RQ1_text$y <- RQ1_text$M - mean(RQ1_text$M)
RQ1_text$y <- RQ1_text$y*1.8 + RQ1_text$M

RQ1_regions <- data %>%
  filter(sex == "Both sexes") %>%
  group_by(agegroup, region, year) %>%
  summarize(M = mean(prob),
            SD = sd(prob),
            N = n(),
            SE = SD/sqrt(N),
            CI = SE*qt(.025, N-1, lower.tail = FALSE))

RQ1_regions_text <- RQ1_regions[RQ1_regions$agegroup == "85+ years",]
RQ1_regions_text <- RQ1_regions_text %>%
  group_by(region) %>%
  mutate(y = seq(10, 40, length.out = 5))

pos <- position_dodge(width = 1)
ggplot(RQ1_regions, aes(x  = agegroup, y = M, group = year, color = year)) +
  geom_line(position = pos, lwd = 1)+
  geom_errorbar(aes(ymin = M - CI, ymax = M + CI), width = 0, position = pos)+
  geom_text(data = RQ1_regions_text[RQ1_regions_text$region == "Africa",], 
             aes(label = year, x = 1, y = y + 20),
             hjust = 0,
             fontface = "bold")+
  geom_text(data = RQ1_regions_text %>% group_by(region) %>% slice(1), 
             aes(label = sprintf("italic(N) == %d", N), x = 15, y = 95, group = region),
             parse = TRUE)+
  facet_wrap(~region)+
  labs(title = "Recent generations live longer across Regions.",
       x = "Age Group",
       y = "Percentage alive",
       color = "Sample Year",
       caption = "Data from the WHO\nErrorbars represent confidence intervals.")+
  scale_y_continuous(limits = c(NA, 100), labels = function(y) paste0(y, " %"))+
  scale_color_viridis_c(option = "plasma", 
                        guide = guide_colourbar(direction = "horizontal", 
                                                title.position = "top", 
                                                title.hjust = 0.5), 
                        end = .8)+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none",
        legend.key.width = unit(3, 'cm'),
        legend.key.height = unit(0.7, 'cm'),
        )

#https://www.datalorax.com/post/alluvial-plots-with-ggforce/

# RQ1_alluv <- RQ1 %>%
#   filter(year == 2019) %>%
#   mutate(dead = 100 - M)
# 
# RQ1_alluv <- RQ1_alluv %>% pivot_longer(cols = c(M, dead))
# RQ1_alluv %>% pivot_wider(names_from = agegroup, values_from = value, id_cols = name)
# 
# library(ggforce)
# 
# gather_set_data(ready, 2:20) %>%
#   ggplot(aes(x = x, id = id, split = y, value = ))

# 
# https://www.alexcernat.com/visualizing-transitions-in-time-using-r/
# RQ1_alluv <- RQ1 %>%
#   filter(year == 2019) %>%
#   mutate(dead = 100 - M) %>% 
#   pivot_longer(cols = c(M, dead))
# 
# ggplot(RQ1_alluv, aes(x = agegroup, y = value, stratum = name, alluvium = name))+
#   geom_stratum()
#   geom_flow()
# 

#####################
```

# Research Question 2: Which countries have the highest survival rate for the age group 85 +?

```{r}
#####################
#Research Question 2#
#####################

#get map data
map <- map_data("world")
map$ISO <- countrycode(map$region, "country.name", "iso3c")

#data that would correspond to research question 2
RQ2 <- data %>%
  filter(sex == "Both sexes" & agegroup == "85+ years" & year == 2019) %>%
  group_by(country, ISO) %>%
  summarize(M = mean(prob))

minmax <- RQ2 %>%
  ungroup() %>%
  arrange(M) %>%
  slice(1:5, (n() - 4) : n())

minmax_countries <- minmax$country

RQ2 <- left_join(map, RQ2)

ggplot(RQ2, aes(x = long, y = lat, group = group, fill = M))+
  geom_polygon(color = "black")+
  scale_fill_viridis_c(option = "plasma",
                       guide = guide_colourbar(title.position = "top", 
                                               title.hjust = 0.5),
                       labels = function(y) paste0(y, " %"))+
  labs(title = "The survival rate at 85 + years varies greatly by country.",
       fill = "Survival Rate at 85+ years",
       caption = "Data from the 2019 Global Health Observatory study.")+
  theme_classic()+
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        axis.line = element_blank(),
        legend.position = "bottom",
        legend.key.width = unit(3, 'cm'),
        legend.key.height = unit(0.7, 'cm'))

```

# Research Question 3: Which countries have the highest survival rate difference for Women vs. Men?
## As a map
```{r}
#####################
#Research Question 3#
#####################

#data that would correspond to research question 3
RQ3 <- data %>%
  filter(sex != "Both sexes" & agegroup == "85+ years" & year == 2019) %>%
  group_by(country, ISO) %>%
  summarize(age_diff = prob[sex == "Female"] - prob[sex == "Male"])

minmax <- RQ3 %>%
  ungroup() %>%
  arrange(age_diff) %>%
  slice(1:5, (n() - 4) : n())

minmax_countries <- minmax$country

RQ3 <- left_join(map, RQ3)

breaks <- (c(min(RQ3$age_diff, na.rm = TRUE), max(RQ3$age_diff, na.rm = TRUE))) %>% round(2)
steps <- seq(breaks[1], breaks[2], by = 1) %>% round() 
steps <- steps[steps %% 5 == 0]
steps <- steps[c(-1, -length(steps))]
breaks <- c(breaks, steps) %>% sort()

ggplot(RQ3, aes(x = long, y = lat, group = group, fill = age_diff))+
  geom_polygon(color = "black")+
  scale_fill_gradient2(low = "blue", mid = "white", midpoint = 0, high = "red", 
                       labels = function(y) paste0(y, " %"),
                       guide = guide_colourbar(title.position = "top", 
                                               title.hjust = 0.5,
                                               ticks.colour = "black",
                                               frame.colour = "black"),
                       limits = c(min(breaks), max(breaks)),
                       breaks = breaks)+
  # scale_fill_viridis_c(option = "plasma",
  #                      guide = guide_colourbar(title.position = "top", 
  #                                              title.hjust = 0.5),
  #                      labels = function(y) paste0(y, " %"))+
  labs(title = "In some countries, women have a greater survival chance at 85 + years.",
       fill = "Difference of survival \nPositive numbers indicate a higher portion of women surviving.")+
  theme_classic()+
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        axis.line = element_blank(),
        legend.position = "bottom",
        legend.key.width = unit(3, 'cm'),
        legend.key.height = unit(0.7, 'cm'))
```

## As a bar chart of the most extreme 10.
### Version 1
```{r}
show_N <- 10
RQ3_bar <- RQ3 %>% 
  group_by(country) %>% 
  slice(1) %>%
  ungroup() %>%
  mutate(country = factor(country),
         country = fct_reorder(country, age_diff, .desc = TRUE)) %>%
  filter(!is.na(age_diff)) %>%
  arrange(age_diff) %>%
  slice(1:show_N, (n() - (show_N-1)) : n())

ggplot(RQ3_bar)+
  geom_bar(aes(x = country, y = age_diff, fill = age_diff), stat = "identity", show.legend = FALSE)+
  scale_fill_viridis_c()+
  geom_vline(xintercept = show_N + 0.5)+
  geom_hline(yintercept = 0, lwd = 1)+
  scale_y_continuous(labels = function(y) paste0(y, " %"))+
  geom_text(aes(y = -10, x = show_N+2, label = "Bottom 10 Countries\nin which women\nhave a higher survival rate"))+
  geom_text(aes(y = -10, x = show_N-1, label = "Top 10 Countries\nin which women\nhave a higher survival rate"))+
  labs(title = "In most countries, women survive longer than men.",
       subtitle = "The most extreme 20 countries worldwide are shown.",
       y = "Difference in survival rate between Women and Men")+
  coord_flip()+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Version 2
```{r}
RQ3_bar2 <- RQ3_bar %>%
  mutate(highlow = rep(c("low", "high"), each = show_N))
  
(high <- ggplot(RQ3_bar2 %>% filter(highlow == "high"))+
  geom_bar(aes(x = country, y = age_diff, fill = age_diff), stat = "identity", show.legend = FALSE)+
  geom_hline(yintercept = 0)+
  scale_fill_viridis_c()+
  scale_y_continuous(limits = c(min(breaks), max(breaks)),
                     labels = function(y) paste0(y, " %"))+
  labs(title = "Bottom 10 Countries\nin which women have a higher survival rate")+
  coord_flip()+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.y = element_blank()))

low <- ggplot(RQ3_bar2 %>% filter(highlow == "low"))+
  geom_bar(aes(x = country, y = age_diff, fill = age_diff), stat = "identity", show.legend = FALSE)+
  geom_hline(yintercept = 0)+
  scale_fill_viridis_c()+
  scale_y_continuous(limits = c(min(breaks), max(breaks)),
                     labels = function(y) paste0(y, " %"))+
  labs(title = "Top 10 Countries\nin which women have a higher survival rate")+
  coord_flip()+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.y = element_blank())

ggarrange(low, high, nrow = 2, common.legend = TRUE)
```


# Bonus Figure: Demographic Development

```{r}
#######
#BONUS#
#######

#This is for the replication of the figure found here: https://upload.wikimedia.org/wikipedia/commons/f/f6/Global_Population-Pyramid-1950-to-2100.jpg

# male_est   <- read.xlsx("data/UN_POPULATION_SINGLE_AGE_MALE.xlsx", sheet = 1, startRow = 17, colNames = TRUE)
# male_pro   <- read.xlsx("data/UN_POPULATION_SINGLE_AGE_MALE.xlsx", sheet = 2, startRow = 17, colNames = TRUE)
# female_est <- read.xlsx("data/UN_POPULATION_SINGLE_AGE_FEMALE.xlsx", sheet = 1, startRow = 17, colNames = TRUE)
# female_pro <- read.xlsx("data/UN_POPULATION_SINGLE_AGE_FEMALE.xlsx", sheet = 2, startRow = 17, colNames = TRUE)
# 
# 
# male_est <- male_est %>%
#   clean_names() %>%
#   filter(region_subregion_country_or_area == "WORLD") %>%
#   select(-index, -variant, -notes, -location_code, -iso2_alpha_code, -iso3_alpha_code, -sdmx_code, -type, -parent_code) %>%
#   mutate(sex = "male",
#          variant = "estimate", .after = region_subregion_country_or_area)
# 
# male_pro <- male_pro %>%
#   clean_names() %>%
#   filter(region_subregion_country_or_area == "WORLD") %>%
#   select(-index, -variant, -notes, -location_code, -iso2_alpha_code, -iso3_alpha_code, -sdmx_code, -type, -parent_code) %>%
#   mutate(sex = "male",
#          variant = "pro", .after = region_subregion_country_or_area)
# 
# female_est <- female_est %>%
#   clean_names() %>%
#   filter(region_subregion_country_or_area == "WORLD") %>%
#   select(-index, -variant, -notes, -location_code, -iso2_alpha_code, -iso3_alpha_code, -sdmx_code, -type, -parent_code) %>%
#   mutate(sex = "female",
#          variant = "estimate", .after = region_subregion_country_or_area)
# 
# female_pro <- female_pro %>%
#   clean_names() %>%
#   filter(region_subregion_country_or_area == "WORLD") %>%
#   select(-index, -variant, -notes, -location_code, -iso2_alpha_code, -iso3_alpha_code, -sdmx_code, -type, -parent_code) %>%
#   mutate(sex = "female",
#          variant = "pro", .after = region_subregion_country_or_area)
# 
# 
# bonus_data <- bind_rows(male_est, male_pro, female_est, female_pro)
# 
# saveRDS(bonus_data, "data/bonus_data.RDS")
bonus_data <- readRDS("data/bonus_data.RDS")

bonus_data2 <- bonus_data %>%
  pivot_longer(cols = starts_with("x"), names_to = "age") %>%
  mutate(age = gsub("x", "", age),
         age = as.numeric(as.character(age)),
         value = as.numeric(value),
         year = as.numeric(year))

bonus_data2$value <- ifelse(bonus_data2$sex == "male", bonus_data2$value*-1, bonus_data2$value)

years_highlighted <- seq(1950, 2100, by = 25)
years_highlighted <- c(1950, 1960, 1970, 1980, 1990, 2018, 2050, 2075, 2100)

bonus_data3 <- bonus_data2[bonus_data2$year %in% years_highlighted,]

helper <- bonus_data3 %>%
  group_by(region_subregion_country_or_area, sex, variant, year) %>%
  summarize(value = 0, age = 0)

helper2 <- bonus_data3 %>%
  group_by(region_subregion_country_or_area, sex, variant, year) %>%
  summarize(value = 0, age = 100)

helper <- bind_rows(helper2, helper, bonus_data3)

helper <- helper %>%
  mutate(year = factor(year, levels = rev(years_highlighted)))

helper_text <- bonus_data3 %>%
  filter(sex == "male") %>%
  group_by(year) %>%
  mutate(index = 1:n(),
         val = abs(value + 35000),
         pos = which.min(val)) %>%
  slice(pos[1])

ggplot()+
  geom_hline(yintercept = seq(10, 100, 10), color = "lightgrey")+
  geom_polygon(data = helper, aes(x = value, y = age, fill = as.numeric(as.character(year)), group = interaction(year, sex)), show.legend = FALSE)+
  geom_path(data = bonus_data3, aes(x = value, y = age, group = interaction(sex, year), linetype = variant), color = "black", lwd = .2)+
  geom_path(data = bonus_data2, aes(x = value, y = age, group = interaction(sex, year), linetype = variant), color = "lightgrey", lwd = .6, alpha = .1)+
  geom_text(data = helper_text, aes(x = -35000, y = age, label = year), angle = 35, color = "black", position = position_nudge(y = +2), size = 3.5)+
  geom_text(aes(x = c(-15000, 15000), y = c(10, 10), label = c("Male", "Female")))+
  geom_vline(xintercept = 0)+
  scale_x_continuous(position = "top", 
                     labels = function(y) paste0(abs(y)/1000, " million"),
                     breaks = seq(-80000, 80000, by = 20000))+
  scale_y_continuous(breaks = seq(10, 100, 10), labels = function(y) paste0(y, " years"))+
  scale_fill_viridis_c(begin = .2, option = "plasma")+
  scale_linetype_discrete(labels = c("estimate" = "estimation", "pro" = "projection"))+
  labs(linetype = "",
       title = "Development of the Worldwide Age Distribution.",
       caption = "Inspired by\n https://upload.wikimedia.org/wikipedia/commons/f/f6/Global_Population-Pyramid-1950-to-2100.jpg",
       y = "Age")+
  theme_classic()+
  theme(legend.position = "bottom")


# ggplot()+
#   geom_smooth(data = bonus_data3, aes(x = value, y = age, color = "grey", group = interaction(year, sex)), se = FALSE, span = .5)+
#   geom_ribbon(data = bonus_data3, aes(x = value, ymin = 0, ymax = predict(loess(age ~ value, span = .5)), group = interaction(year, sex), fill = as.numeric(as.character(year))),
#                 alpha = 0.3)+
#   scale_fill_viridis_c(begin = .2, option = "plasma")+
#   theme_classic()
# 
# 
# 
# ggplot()+
#   geom_line(data = bonus_data3, aes(x = value, y = predict(loess(age ~ value, span = .5)), color = "grey", group = interaction(year, sex)))+
#   scale_fill_viridis_c(begin = .2, option = "plasma")+
#   theme_classic()
# 
# 
# p <- ggplot()
# 
# for(i in 1:length(years_highlighted)){
#   p <- p +
#     geom_line(data = bonus_data3 %>% filter(year == years_highlighted[i],
#                                               sex == "male"), 
#                 aes(x = value, 
#                     y = age, 
#                     group = interaction(year, sex)), 
#                 se = FALSE, span = .5)+
#     geom_ribbon(data = bonus_data3 %>% filter(year == years_highlighted[i],
#                                               sex == "male"), 
#                 aes(x = value, 
#                     ymin = 0, 
#                     ymax = predict(loess(age ~ value, span = .5)), 
#                     group = interaction(year, sex), 
#                     fill = as.numeric(as.character(year))),
#                 alpha = 0.3)+
#     geom_smooth(data = bonus_data3 %>% filter(year == years_highlighted[i],
#                                               sex == "female"), 
#                 aes(x = value, 
#                     y = age, 
#                     group = interaction(year, sex)), 
#                 se = FALSE, span = .5)+
#     geom_ribbon(data = bonus_data3 %>% filter(year == years_highlighted[i],
#                                               sex == "female"), 
#                 aes(x = value, 
#                     ymin = 0, 
#                     ymax = predict(loess(age ~ value, span = .5)), 
#                     group = interaction(year, sex), 
#                     fill = as.numeric(as.character(year))),
#                 alpha = 0.3)
#   
# }
# 
# p +
#   scale_fill_viridis_c(begin = .2, option = "plasma")+
#   theme_classic()
#   

```
