---
title: "Final Project"
author: "Dominik Graetz"
date: "2023-02-25"
output: 
  html_document:
    toc: true
    toc_float: true
---

This proposal works with data from the World Health Organization and United Nations. All of the research questions are related to demographics, more specifically the development of the survival rate within age groups. The [dataset from the WHO](https://www.who.int/data/gho/indicator-metadata-registry/imr-details/31) which will be used for the main three research questions has survival data for 183 countries, separated out by sex (male, female, combined), 19 age groups and 5 waves. The dataset can be downloaded using the `{WHO}` package, but is read in as an .RDS because this is much faster and works more reliably.

I will be looking at the following Research questions:

1) Across countries, did the survival rate generally increase from 2000 to 2019 across countries?

2) Which countries have the highest survival rate for the age group of 85+ years?

3) Which countries show the most extreme survival rate differences between Women and Men?

Furthermore, I came across this cool [graphic](https://upload.wikimedia.org/wikipedia/commons/f/f6/Global_Population-Pyramid-1950-to-2100.jpg) that I will be trying to reproduce. This graphic uses data from the UN which can be found [here](https://population.un.org/wpp/Download/Standard/Population/). However, since the files are too big (> 100 MB), they cannot be hosted on github. Instead, the data set necessary for this graphic was generated from the original files and saved as an RDS. It is much smaller in size and is hosted on github. 

# Initializing, Data loading
This code takes forever, so I executed it only once and saved the data as an RDS
```{r setup, include=FALSE}

knitr::opts_chunk$set(cache = TRUE, dev = "svg")
#library(WHO)
#y <- get_data("LIFE_0000000031")
#saveRDS(y, "data/WHO_survival_data.RDS")

options(scipen = 999)
library(tidyverse)
library(forcats)
library(janitor)
library(openxlsx)
library(ggpubr)
library(countrycode)

data <- readRDS("data/WHO_survival_data.RDS")
countries <- read.csv("data/WHO_country_conversion.csv")

data <- data %>%
  mutate(agegroup = factor(agegroup),
         agegroup = factor(agegroup, levels = c(levels(agegroup)[1:2], levels(agegroup)[11], levels(agegroup)[3:10], levels(agegroup)[12:19])),
         prob = (value/100000)*100) %>%
  filter(!is.na(country),
         !is.na(agegroup))

data <- left_join(data, countries %>% select(DisplayString, ISO), by = c("country" = "DisplayString"))

```

# Exploratory Plot

```{r}

#useless plot, plotting individual survival traces for all countries for 2019
ggplot(data[data$year == 2019 & data$sex != "Both sexes",])+
  geom_line(aes(x = agegroup, y = prob, group = interaction(country, sex), color = sex), alpha = .2)+
  labs(caption = "Data from the WHO, https://www.who.int/data/gho/indicator-metadata-registry/imr-details/31")+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1))
```

# Research Question 1: Across countries, did the survival rate generally increase from 2000 to 2019 across countries?

## Attempt 1, very simple
```{r}
#May be worth splitting by Region.
RQ1 <- data %>%
  filter(sex == "Both sexes") %>%
  group_by(agegroup, year) %>%
  summarize(M = mean(prob),
            SD = sd(prob),
            N = n(),
            SE = SD/sqrt(N),
            CI = SE*qt(.025, N-1, lower.tail = FALSE))

ggplot(RQ1) +
  geom_line(aes(x  = agegroup, y = M, group = year, color = year))+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1))

```

## Errorbars!

```{r}
pos <- position_dodge(width = .6)
ggplot(RQ1, aes(x  = agegroup, y = M, group = year, color = year)) +
  geom_line(position = pos, lwd = 1)+
  scale_color_continuous(breaks = c(2000, 2005, 2010, 2015, 2019))+
  geom_errorbar(aes(ymin = M - CI, ymax = M + CI), width = 0, position = pos, lwd = 1)+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## That's a nice first version!
Thinking about removing the color bar and just leaving the year as text. 

```{r}
RQ1_text <- RQ1[RQ1$agegroup == "85+ years",]
RQ1_text$y <- RQ1_text$M - mean(RQ1_text$M)
RQ1_text$y <- RQ1_text$y*1.5 + RQ1_text$M

pos <- position_dodge(width = .6)
ggplot(RQ1, aes(x  = agegroup, y = M, group = year, color = year)) +
  geom_line(position = pos, lwd = 1)+
  geom_errorbar(aes(ymin = M - CI, ymax = M + CI), width = 0, position = pos, lwd = 1)+
  geom_text(data = RQ1_text, aes(label = year, y = y), position = pos, hjust = -0.4, fontface = "bold")+
  labs(title = "Younger generations live longer.",
       subtitle = paste0("Data averaged across ", RQ1$N[1], " countries"),
       x = "Age Group",
       y = "Percentage alive",
       color = "Sample Year",
       caption = "Errorbars represent confidence intervals.\n Data from the WHO, https://www.who.int/data/gho/indicator-metadata-registry/imr-details/31")+
  scale_y_continuous(limits = c(0, 100), 
                     labels = function(y) paste0(y, " %"))+
  scale_color_viridis_c(option = "plasma", 
                        guide = guide_colourbar(direction = "horizontal", 
                                                title.position = "top"), 
                        end = .9,
                        breaks = c(2000, 2005, 2010, 2015, 2019))+
  expand_limits(x = c(0, 22))+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = c(0.1, 0.2),
        legend.key.width = unit(1, 'cm'),
        legend.key.height = unit(0.4, 'cm'),
        legend.justification = "left")

```

## By region
Probably worth it, there are some interesting differences.

```{r}
#BY REGION

RQ1_text <- RQ1[RQ1$agegroup == "85+ years",]
RQ1_text$y <- RQ1_text$M - mean(RQ1_text$M)
RQ1_text$y <- RQ1_text$y*1.8 + RQ1_text$M

RQ1_regions <- data %>%
  filter(sex == "Both sexes") %>%
  group_by(agegroup, region, year) %>%
  summarize(M = mean(prob),
            SD = sd(prob),
            N = n(),
            SE = SD/sqrt(N),
            CI = SE*qt(.025, N-1, lower.tail = FALSE))

RQ1_regions_text <- RQ1_regions[RQ1_regions$agegroup == "85+ years",]
RQ1_regions_text <- RQ1_regions_text %>%
  group_by(region) %>%
  mutate(y = seq(10, 40, length.out = 5))

pos <- position_dodge(width = 1)
ggplot(RQ1_regions, aes(x  = agegroup, y = M, group = year, color = year)) +
  geom_line(position = pos, lwd = 1)+
  geom_errorbar(aes(ymin = M - CI, ymax = M + CI), width = 0, position = pos)+
  geom_text(data = RQ1_regions_text[RQ1_regions_text$region == "Africa",], 
             aes(label = year, x = 1, y = y + 20),
             hjust = 0,
             fontface = "bold")+
  geom_text(data = RQ1_regions_text %>% group_by(region) %>% slice(1), 
             aes(label = sprintf("italic(N) == %d", N), x = 15, y = 95, group = region),
             parse = TRUE)+
  facet_wrap(~region)+
  labs(title = "Recent generations live longer across Regions.",
       x = "Age Group",
       y = "Percentage alive",
       color = "Sample Year",
       caption = "Errorbars represent confidence intervals.\n Data from the WHO, https://www.who.int/data/gho/indicator-metadata-registry/imr-details/31")+
  scale_y_continuous(limits = c(NA, 100), labels = function(y) paste0(y, " %"))+
  scale_color_viridis_c(option = "plasma", 
                        guide = guide_colourbar(direction = "horizontal", 
                                                title.position = "top", 
                                                title.hjust = 0.5), 
                        end = .8)+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none",
        legend.key.width = unit(3, 'cm'),
        legend.key.height = unit(0.7, 'cm'),
        )

#https://www.datalorax.com/post/alluvial-plots-with-ggforce/

# RQ1_alluv <- RQ1 %>%
#   filter(year == 2019) %>%
#   mutate(dead = 100 - M)
# 
# RQ1_alluv <- RQ1_alluv %>% pivot_longer(cols = c(M, dead))
# RQ1_alluv %>% pivot_wider(names_from = agegroup, values_from = value, id_cols = name)
# 
# library(ggforce)
# 
# gather_set_data(ready, 2:20) %>%
#   ggplot(aes(x = x, id = id, split = y, value = ))

# 
# https://www.alexcernat.com/visualizing-transitions-in-time-using-r/
# RQ1_alluv <- RQ1 %>%
#   filter(year == 2019) %>%
#   mutate(dead = 100 - M) %>% 
#   pivot_longer(cols = c(M, dead))
# 
# ggplot(RQ1_alluv, aes(x = agegroup, y = value, stratum = name, alluvium = name))+
#   geom_stratum()
#   geom_flow()
# 

#####################
```

# Research Question 2: Which countries have the highest survival rate for the age group of 85+ years?

```{r}
#####################
#Research Question 2#
#####################

#get map data
map <- map_data("world")
map$ISO <- countrycode(map$region, "country.name", "iso3c")

#data that would correspond to research question 2
RQ2 <- data %>%
  filter(sex == "Both sexes" & agegroup == "85+ years" & year == 2019) %>%
  group_by(country, ISO) %>%
  summarize(M = mean(prob))

minmax <- RQ2 %>%
  ungroup() %>%
  arrange(M) %>%
  slice(1:5, (n() - 4) : n())

minmax_countries <- minmax$country

RQ2 <- left_join(map, RQ2)

ggplot(RQ2, aes(x = long, y = lat, group = group, fill = M))+
  geom_polygon(color = "black")+
  scale_fill_viridis_c(option = "plasma",
                       guide = guide_colourbar(title.position = "top", 
                                               title.hjust = 0.5),
                       labels = function(y) paste0(y, " %"))+
  labs(title = "The survival rate at 85 + years varies greatly by country.",
       fill = "Survival Rate at 85+ years",
       caption = "Data from the WHO, https://www.who.int/data/gho/indicator-metadata-registry/imr-details/31")+
  theme_classic()+
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        axis.line = element_blank(),
        legend.position = "bottom",
        legend.key.width = unit(3, 'cm'),
        legend.key.height = unit(0.7, 'cm'))

```

# Research Question 3: Which countries show the most extreme survival rate differences between Women and Men?
## As a map
```{r}
#####################
#Research Question 3#
#####################

#data that would correspond to research question 3
RQ3 <- data %>%
  filter(sex != "Both sexes" & agegroup == "85+ years" & year == 2019) %>%
  group_by(country, ISO) %>%
  summarize(age_diff = prob[sex == "Female"] - prob[sex == "Male"])

minmax <- RQ3 %>%
  ungroup() %>%
  arrange(age_diff) %>%
  slice(1:5, (n() - 4) : n())

minmax_countries <- minmax$country

RQ3 <- left_join(map, RQ3)

breaks <- (c(min(RQ3$age_diff, na.rm = TRUE), max(RQ3$age_diff, na.rm = TRUE))) %>% round(2)
steps <- seq(breaks[1], breaks[2], by = 1) %>% round() 
steps <- steps[steps %% 5 == 0]
steps <- steps[c(-1, -length(steps))]
breaks <- c(breaks, steps) %>% sort()

ggplot(RQ3, aes(x = long, y = lat, group = group, fill = age_diff))+
  geom_polygon(color = "black")+
  scale_fill_gradient2(low = "blue", mid = "white", midpoint = 0, high = "red", 
                       labels = function(y) paste0(y, " %"),
                       guide = guide_colourbar(title.position = "top", 
                                               title.hjust = 0.5,
                                               ticks.colour = "black",
                                               frame.colour = "black"),
                       limits = c(min(breaks), max(breaks)),
                       breaks = breaks)+
  # scale_fill_viridis_c(option = "plasma",
  #                      guide = guide_colourbar(title.position = "top", 
  #                                              title.hjust = 0.5),
  #                      labels = function(y) paste0(y, " %"))+
  labs(title = "In some countries, women have a greater survival chance at 85 + years.",
       fill = "Difference of survival \nPositive numbers indicate a higher portion of women surviving.",
       caption = "Data from the WHO, https://www.who.int/data/gho/indicator-metadata-registry/imr-details/31")+
  theme_classic()+
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        axis.line = element_blank(),
        legend.position = "bottom",
        legend.key.width = unit(3, 'cm'),
        legend.key.height = unit(0.7, 'cm'))
```

## As a bar chart of the most extreme 10.
### Version 1
```{r}
show_N <- 10
RQ3_bar <- RQ3 %>% 
  group_by(country) %>% 
  slice(1) %>%
  ungroup() %>%
  mutate(country = factor(country),
         country = fct_reorder(country, age_diff, .desc = TRUE)) %>%
  filter(!is.na(age_diff)) %>%
  arrange(age_diff) %>%
  slice(1:show_N, (n() - (show_N-1)) : n())

ggplot(RQ3_bar)+
  geom_bar(aes(x = country, y = age_diff, fill = age_diff), stat = "identity", show.legend = FALSE)+
  scale_fill_viridis_c()+
  geom_vline(xintercept = show_N + 0.5)+
  geom_hline(yintercept = 0, lwd = 1)+
  scale_y_continuous(labels = function(y) paste0(y, " %"))+
  geom_text(aes(y = -10, x = show_N+2, label = "Bottom 10 Countries\nin which women\nhave a higher survival rate"))+
  geom_text(aes(y = -10, x = show_N-1, label = "Top 10 Countries\nin which women\nhave a higher survival rate"))+
  labs(title = "In most countries, women survive longer than men.",
       subtitle = "The most extreme 20 countries worldwide are shown.",
       y = "Difference in survival rate between Women and Men",
       caption = "Data from the WHO, https://www.who.int/data/gho/indicator-metadata-registry/imr-details/31")+
  coord_flip()+
  theme_classic()
```

### Version 2
```{r}
RQ3_bar2 <- RQ3_bar %>%
  mutate(highlow = rep(c("low", "high"), each = show_N))
  
high <- ggplot(RQ3_bar2 %>% filter(highlow == "high"))+
  geom_bar(aes(x = country, y = age_diff, fill = age_diff), stat = "identity", show.legend = FALSE)+
  geom_hline(yintercept = 0)+
  scale_fill_viridis_c()+
  scale_y_continuous(limits = c(min(breaks), max(breaks)),
                     labels = function(y) paste0(y, " %"))+
  labs(title = "Bottom 10 Countries\nin which women have a higher survival rate",
       caption = "Data from the WHO, https://www.who.int/data/gho/indicator-metadata-registry/imr-details/31")+
  coord_flip()+
  theme_classic()+
  theme(axis.title.y = element_blank())

low <- ggplot(RQ3_bar2 %>% filter(highlow == "low"))+
  geom_bar(aes(x = country, y = age_diff, fill = age_diff), stat = "identity", show.legend = FALSE)+
  geom_hline(yintercept = 0)+
  scale_fill_viridis_c()+
  scale_y_continuous(limits = c(min(breaks), max(breaks)),
                     labels = function(y) paste0(y, " %"))+
  labs(title = "Top 10 Countries\nin which women have a higher survival rate")+
  coord_flip()+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.y = element_blank())

ggarrange(low, high, nrow = 2, common.legend = TRUE)
```


# Bonus Figure: Demographic Development

```{r}
#######
#BONUS#
#######

#This section reproduces the figure found here: https://upload.wikimedia.org/wikipedia/commons/f/f6/Global_Population-Pyramid-1950-to-2100.jpg
# Data from https://population.un.org/wpp/Download/Standard/Population/

#If you want to download the data within R, you need to uncomment and run the following code once:

#Download data - these are fairly large (> 100 MB each and take a bit of time)
# options(timeout = 500)
# 
# download.file(url = "https://population.un.org/wpp/Download/Files/1_Indicators%20(Standard)/EXCEL_FILES/2_Population/WPP2022_POP_F01_3_POPULATION_SINGLE_AGE_FEMALE.xlsx", destfile = "data/UN_Female.xlsx", mode = "wb")
# 
# download.file(url = "https://population.un.org/wpp/Download/Files/1_Indicators%20(Standard)/EXCEL_FILES/2_Population/WPP2022_POP_F01_2_POPULATION_SINGLE_AGE_MALE.xlsx", destfile = "data/UN_Male.xlsx", mode = "wb")
# 
# male_est   <- read.xlsx("data/UN_Male.xlsx", sheet = 1, startRow = 17, colNames = TRUE)
# male_pro   <- read.xlsx("data/UN_Male.xlsx", sheet = 2, startRow = 17, colNames = TRUE)
# female_est <- read.xlsx("data/UN_Female.xlsx", sheet = 1, startRow = 17, colNames = TRUE)
# female_pro <- read.xlsx("data/UN_Female.xlsx", sheet = 2, startRow = 17, colNames = TRUE)
# 
# 
# male_est <- male_est %>%
#   clean_names() %>%
#   filter(region_subregion_country_or_area == "WORLD") %>%
#   select(-index, -variant, -notes, -location_code, -iso2_alpha_code, -iso3_alpha_code, -sdmx_code, -type, -parent_code) %>%
#   mutate(sex = "male",
#          variant = "estimate", .after = region_subregion_country_or_area)
# 
# male_pro <- male_pro %>%
#   clean_names() %>%
#   filter(region_subregion_country_or_area == "WORLD") %>%
#   select(-index, -variant, -notes, -location_code, -iso2_alpha_code, -iso3_alpha_code, -sdmx_code, -type, -parent_code) %>%
#   mutate(sex = "male",
#          variant = "pro", .after = region_subregion_country_or_area)
# 
# female_est <- female_est %>%
#   clean_names() %>%
#   filter(region_subregion_country_or_area == "WORLD") %>%
#   select(-index, -variant, -notes, -location_code, -iso2_alpha_code, -iso3_alpha_code, -sdmx_code, -type, -parent_code) %>%
#   mutate(sex = "female",
#          variant = "estimate", .after = region_subregion_country_or_area)
# 
# female_pro <- female_pro %>%
#   clean_names() %>%
#   filter(region_subregion_country_or_area == "WORLD") %>%
#   select(-index, -variant, -notes, -location_code, -iso2_alpha_code, -iso3_alpha_code, -sdmx_code, -type, -parent_code) %>%
#   mutate(sex = "female",
#          variant = "pro", .after = region_subregion_country_or_area)
# 
# 
# bonus_data <- bind_rows(male_est, male_pro, female_est, female_pro)
# 
# saveRDS(bonus_data, "data/bonus_data.RDS")
bonus_data <- readRDS("data/bonus_data.RDS")

bonus_data2 <- bonus_data %>%
  pivot_longer(cols = starts_with("x"), names_to = "age") %>%
  mutate(age = gsub("x", "", age),
         age = as.numeric(as.character(age)),
         value = as.numeric(value),
         year = as.numeric(year))

bonus_data2$value <- ifelse(bonus_data2$sex == "male", bonus_data2$value*-1, bonus_data2$value)

years_highlighted <- seq(1950, 2100, by = 25)
years_highlighted <- c(1950, 1960, 1970, 1980, 1990, 2018, 2050, 2075, 2100)

bonus_data3 <- bonus_data2[bonus_data2$year %in% years_highlighted,]

helper <- bonus_data3 %>%
  group_by(region_subregion_country_or_area, sex, variant, year) %>%
  summarize(value = 0, age = 0)

helper2 <- bonus_data3 %>%
  group_by(region_subregion_country_or_area, sex, variant, year) %>%
  summarize(value = 0, age = 100)

helper <- bind_rows(helper2, helper, bonus_data3)

helper <- helper %>%
  mutate(year = factor(year, levels = rev(years_highlighted)))

helper_text <- bonus_data3 %>%
  filter(sex == "male") %>%
  group_by(year) %>%
  mutate(index = 1:n(),
         val = abs(value + 35000),
         pos = which.min(val)) %>%
  slice(pos[1])

ggplot()+
  geom_hline(yintercept = seq(10, 100, 10), color = "lightgrey")+
  geom_polygon(data = helper, aes(x = value, y = age, fill = as.numeric(as.character(year)), group = interaction(year, sex)), show.legend = FALSE)+
  geom_path(data = bonus_data3, aes(x = value, y = age, group = interaction(sex, year), linetype = variant), color = "black", lwd = .2)+
  geom_path(data = bonus_data2, aes(x = value, y = age, group = interaction(sex, year), linetype = variant), color = "lightgrey", lwd = .6, alpha = .1)+
  geom_text(data = helper_text, aes(x = -35000, y = age, label = year), angle = 35, color = "black", position = position_nudge(y = +2), size = 3.5)+
  geom_text(aes(x = c(-15000, 15000), y = c(10, 10), label = c("Male", "Female")))+
  geom_vline(xintercept = 0)+
  scale_x_continuous(position = "top", 
                     labels = function(y) paste0(abs(y)/1000, " million"),
                     breaks = seq(-80000, 80000, by = 20000))+
  scale_y_continuous(breaks = seq(10, 100, 10), labels = function(y) paste0(y, " years"))+
  scale_fill_viridis_c(begin = .2, option = "plasma")+
  scale_linetype_discrete(labels = c("estimate" = "estimation", "pro" = "projection"))+
  labs(linetype = "",
       title = "Development of the Worldwide Age Distribution.",
       caption = "Inspired by\n https://upload.wikimedia.org/wikipedia/commons/f/f6/Global_Population-Pyramid-1950-to-2100.jpg\n Data from United Nations, https://population.un.org/wpp/Download/Standard/Population/",
       y = "Age")+
  theme_classic()+
  theme(legend.position = "bottom")


#This is work in progress:
# ggplot()+
#   geom_smooth(data = bonus_data3, aes(x = value, y = age, color = "grey", group = interaction(year, sex)), se = FALSE, span = .5)+
#   geom_ribbon(data = bonus_data3, aes(x = value, ymin = 0, ymax = predict(loess(age ~ value, span = .5)), group = interaction(year, sex), fill = as.numeric(as.character(year))),
#                 alpha = 0.3)+
#   scale_fill_viridis_c(begin = .2, option = "plasma")+
#   theme_classic()
# 
# 
# 
# ggplot()+
#   geom_line(data = bonus_data3, aes(x = value, y = predict(loess(age ~ value, span = .5)), color = "grey", group = interaction(year, sex)))+
#   scale_fill_viridis_c(begin = .2, option = "plasma")+
#   theme_classic()
# 
# 
# p <- ggplot()
# 
# for(i in 1:length(years_highlighted)){
#   p <- p +
#     geom_line(data = bonus_data3 %>% filter(year == years_highlighted[i],
#                                               sex == "male"), 
#                 aes(x = value, 
#                     y = age, 
#                     group = interaction(year, sex)), 
#                 se = FALSE, span = .5)+
#     geom_ribbon(data = bonus_data3 %>% filter(year == years_highlighted[i],
#                                               sex == "male"), 
#                 aes(x = value, 
#                     ymin = 0, 
#                     ymax = predict(loess(age ~ value, span = .5)), 
#                     group = interaction(year, sex), 
#                     fill = as.numeric(as.character(year))),
#                 alpha = 0.3)+
#     geom_smooth(data = bonus_data3 %>% filter(year == years_highlighted[i],
#                                               sex == "female"), 
#                 aes(x = value, 
#                     y = age, 
#                     group = interaction(year, sex)), 
#                 se = FALSE, span = .5)+
#     geom_ribbon(data = bonus_data3 %>% filter(year == years_highlighted[i],
#                                               sex == "female"), 
#                 aes(x = value, 
#                     ymin = 0, 
#                     ymax = predict(loess(age ~ value, span = .5)), 
#                     group = interaction(year, sex), 
#                     fill = as.numeric(as.character(year))),
#                 alpha = 0.3)
#   
# }
# 
# p +
#   scale_fill_viridis_c(begin = .2, option = "plasma")+
#   theme_classic()
#   

```


```{r}
sessionInfo()
```

